---
title: "Preterm Birth"
subtitle: "Using Random Forest to Predict Preterm births among women in rural settings of Uganda."
author: "Mercy Consolate Akello"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: hide
    theme: cerulean
    toc: yes
    highlight: breezedark
    toc_depth: 2
---

# Load data
```{r message=FALSE}
library(readr)
Pretem <- read_csv("C:/Users/Akello_Mercy/Desktop/wwcode hackathon/preterm.csv")
Preterm=Pretem
```

# load required packages
```{r message=FALSE}
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(forcats)
library(stringr)
```

## Inspect data
```{r}
glimpse(Preterm)
```

## inspect a few of the observations
```{r}
tail(Preterm)
```

# Clean the data
## create new categorical variable from binary columns
```{r}
Preterm$anc.attendant=as.factor(ifelse(Preterm$Q409__1==1,"Doctor",
                ifelse(Preterm$Q409__2==1,"Nurse",
                ifelse(Preterm$Q409__3==1,"Midwife",
                ifelse(Preterm$Q409__4==1,"Other",
                ifelse(Preterm$Q409__5==1,"Other",
                ifelse(Preterm$Q409__99==1,"Other",NA)))))))

Preterm$place.of.anc=as.factor(ifelse(Preterm$"Q410__1"==1,"Other",
                ifelse(Preterm$"Q410__2"==1,"Other",
                ifelse(Preterm$"Q410__3"==1,"Government Hospital",
                ifelse(Preterm$"Q410__4"==1,"Government Health Center",
                ifelse(Preterm$"Q410__5"==1,"Government Health Center",
                ifelse(Preterm$"Q410__6"==1,"Other Government Health Facility",
                ifelse(Preterm$"Q410__7"==1,"Private Hospital",
                ifelse(Preterm$"Q410__8"==1,"Other",
                ifelse(Preterm$"Q410__9"==1,"Other",
                ifelse(Preterm$"Q410__10"==1,"Other", NA)))))))))))
```

## Look at ethinicity
```{r}
table(Preterm$ethinicity)
```

## recategorize values for ethinicity
```{r}
Preterm=Preterm%>%
  mutate(ethinicity=recode(ethinicity,
                           "Acholis"="Other",
                           "Bagisu"="Other",
                           "Bakiga"="Other",
                           "Banyankole"="Other",
                           "Itesot"="Other",
                           "Jopadhola"="Other",
                           "Karamajong"="Other",
                           "Langi"="Other",
                           "Lugubar"="Other"))
```

## inspect religion
```{r}
table(Preterm$religion)
```
**Other has only one observation, combine it with SDA and call it Other.**

## recategorize values for religion
```{r}
Preterm=Preterm%>%
  mutate(religion=recode(religion,
                           "SDA"="Other"))
```

## remove columns that won't be used in analysis
```{r}
Preterm=Preterm%>%
  select(-c(Q409__1:Q409__99, Q410__1:Q410__10))
```


# check for missing values
```{r}
colSums(is.na(Preterm))
```

## replace NA values in education
```{r}
Preterm=Preterm%>%
  replace_na(list(education="None"))
```

## confirm that there are no more NA values
```{r}
anyNA(Preterm$education)
```

# Perform variable casting
```{r}
Preterm$preterm=as.factor(Preterm$preterm)
Preterm$attended.school=as.factor(Preterm$attended.school)
Preterm$education=as.factor(Preterm$education)
Preterm$religion=as.factor(Preterm$religion)
Preterm$ethinicity=as.factor(Preterm$ethinicity)
Preterm$attended.anc=as.factor(Preterm$attended.anc)
Preterm$place.of.anc=as.factor(Preterm$place.of.anc)
Preterm$anc.attendant=as.factor(Preterm$anc.attendant)
```

## change all values in education to sentence case.
```{r}
Preterm$education=str_to_sentence(Preterm$education)
```

# Visualize data
```{r}
ggplot(Preterm,aes(preterm))+
  geom_bar(fill="#32C7A9")+
  labs(title = 'Preterm births')+
  theme_classic()
```

## education
```{r}
Preterm%>%
  group_by(education)%>%
  summarize(count=n())%>%
ggplot(aes(x = reorder(education,count), y = count))+
  geom_bar(stat="identity", fill="#32C7A9")+
  labs(x='education', title = 'Education level of the mothers')
```

## place.of.anc
```{r}
Preterm%>%
  group_by(place.of.anc)%>%
  summarize(count=n())%>%
ggplot(aes(x = reorder(place.of.anc,count), y = count))+
  geom_bar(stat="identity", fill="#32C7A9")+
  coord_flip()+
  labs(x='Place of ANC', title = 'The place of ANC visits for the mothers')
```
**Most mothers attended ANC at Government Health Center or Government Hospital.**


## anc.attendant
```{r}
Preterm%>%
  group_by(anc.attendant)%>%
  summarize(count=n())%>%
ggplot(aes(x = reorder(anc.attendant,count), y = count))+
  geom_bar(stat="identity", fill="#32C7A9")+
  coord_flip()+
  labs(x='ANC Attendant', title = 'The ANC Attendant of the the mother')
```
**Majority of the mothers were attended to by a Nurse.**

# Numeric data
## mother's age
```{r}
Preterm%>%
  ggplot(aes(mothers.age))+
  geom_histogram(aes(y=after_stat(density)),bins=15, fill="#D60093")+
  geom_density(lwd=1, color="#E9BF35")+
  labs(x='mothers.age', title = 'The mothers age distribution')
```
**The age distribution is right skewed with most mothers between the age of 25 and 35 years.**

## number.of.anc.visits
```{r message=FALSE}
Preterm%>%
  ggplot(aes(number.of.anc.visits))+
  geom_histogram(fill="#D60093")+
  labs(x='Number of ANC visits', title = 'Distribution of the number of ANC visits.')
```
**There are some outliers in the data.**

**Find out how many observations are outliers**
```{r}
table(Preterm$number.of.anc.visits)
```

**Find out if they are preterm or not**
```{r}
table(Preterm$number.of.anc.visits,Preterm$preterm)
```

**The values are all not preterm, so we can just drop them.**
```{r}
Preterm=Preterm%>%
  filter(!number.of.anc.visits>10)
```

**Plot number.of.anc.visits without the outlier**
```{r}
Preterm%>%
  ggplot(aes(number.of.anc.visits))+
  geom_histogram(stat="count", fill="#D60093")+
  labs(x='Number of ANC visits', title = 'Distribution of the number of ANC visits.')
```

**Replace 10 with 9**
```{r}
Preterm$number.of.anc.visits=ifelse(Preterm$number.of.anc.visits==10,9,Preterm$number.of.anc.visits)
```

**Plot number.of.anc.visits again**
```{r warning=FALSE}
Preterm%>%
  ggplot(aes(number.of.anc.visits))+
  geom_histogram(stat="count", fill="#D60093")+
  labs(x='Number of ANC visits', title = 'Distribution of the number of ANC visits.')
```
**The distribution of the number of ANC visits is right skewed and majority of mothers attended atleast 4 ANC visits.**

## mothers.age VS number.of.anc.visits
```{r}
Preterm%>%
  ggplot(aes(mothers.age,number.of.anc.visits))+
  geom_point(color="#E9BF35", size=2)+
  labs(x="Mother's Age", y='Number of ANC visits', title = 'Scatter Plot of Mothers Age and Number of ANC visits.')+
  theme_classic()
```
**There appears to be no linear relationship between mother's age and the number of anc visits she makes.**

#  compare mother's age with other categorical variables
## education
```{r}
Preterm%>%
  ggplot(aes(reorder(education,mothers.age),mothers.age))+
  geom_boxplot(fill="#E9BF35")+
  labs(x="Education", y='Mothers Age', title = 'Boxplot of Mothers Age and Education.')+
  theme_classic()
```

## ethinicity
```{r}
Preterm%>%
  ggplot(aes(reorder(ethinicity,mothers.age,median),mothers.age))+
  geom_boxplot(fill="#E9BF35", outlier.color= "#32C7A9",
               outlier.size =2 )+
   labs(x="Ethinicity", y='Mothers Age', title = 'Boxplot of Mothers Age and Ethinicity')+
  theme_classic()
```

## anc.attendant
```{r}
Preterm%>%
  ggplot(aes(reorder(anc.attendant,mothers.age,median),mothers.age))+
  geom_boxplot(fill="#E9BF35")+
   labs(x="ANC Attendant", y='Mothers Age', title = 'Boxplot of Mothers Age and ANC Attendant')+
  theme_classic()
```

## place.of.anc
```{r}
Preterm%>%
  ggplot(aes(reorder(place.of.anc,mothers.age,median),mothers.age))+
  geom_boxplot(fill="#E9BF35")+
   labs(x="Place of ANC", y='Mothers Age', title = 'Boxplot of Mothers Age and Place of ANC')+
  theme_classic()
```

## preterm
```{r}
Preterm%>%
  ggplot(aes(reorder(preterm,-mothers.age,median),mothers.age))+
  geom_boxplot(fill="#E9BF35")+
   labs(x="Preterm", y='Mothers Age', title = 'Boxplot of Mothers Age and Preterm')+
  theme_classic()
```
**The median age of mothers who had preterm delivery(1) was lower than that of mothers who did not have preterm delivery(0).**

# Build a machine learning model
```{r message=FALSE}
require(randomForest)
require(caret)
```

## Divide into Train and Test datasets
```{r}
set.seed(1234)
pre <- sample(2, nrow(Preterm),replace = TRUE,prob = c(0.7,0.3))
Train <- Preterm[pre==1,]
Test <- Preterm[pre==2,]
```

## Random Forest
```{r}
rf=randomForest(preterm~mothers.age+education+religion+ethinicity+number.of.anc.visits+anc.attendant+place.of.anc, Preterm)
```

## predict the preterm using the Test dataset
```{r}
preterm.predicted=predict(rf,Test)
```

## View Predicted values
```{r}
preterm.predicted
```

## model performance
```{r}
Trf <- table(PredictedValue=preterm.predicted, ActualValue=Test$preterm)
print(confusionMatrix(Trf))
```

# Model improvement
**I will now adjust the parameters of the model and rerun the model and see if there is any improvement.**
```{r}
rf2=randomForest(preterm~mothers.age+education+religion+ethinicity+number.of.anc.visits+anc.attendant+place.of.anc, Preterm, ntree=500, mtry=3, nodesize=0.01*nrow(Test))
```

## predict the preterm using the Test dataset
```{r}
preterm.predicted2=predict(rf2,Test)
```

## view Predicted values
```{r}
preterm.predicted2
```

## model performance
```{r}
Trf2 <- table(PredictedValue=preterm.predicted2,
              ActualValue=Test$preterm)
print(confusionMatrix(Trf2))
```
**There is a slight improvement in the model performance.**

# variable importance
```{r}
varImpPlot(rf2, main = "Variable Importance Plot", col="#32C7A9")
```
**The mother's age was the most important variable in predicting preterm birth;this was followed by religion, education, number.of.anc.visits and ethinicity as the top five predictors of preterm birth.**

**place.of.anc and anc.attendant were least important predictors of preterm delivery.**

# Conclusion
**The Random forest model here gives us the mother's age as a key determinant of preterm birth.**

# Recommendation
**Based on this finding; I suggest that policy makers design programs that are focused on addressing the age specific needs of the mothers in rural communities.**


